Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models

### 2015-08-20 TeXFiles/Manuscript_Albert_Ulzega.tex
FF ref added. ready for Carlo

The methods we present are generic and applicable to a wide range of SDE models.
However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
The model equation is thus given by the SDE
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed (ref \textbf{FF}).

@@

The algorithm described above was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9)~\cite{julia_web}. One of the attractive features of the language is its (fast growing) library of over 600 packages for numerical and scientific applications.
In our case, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions. Our algorithm benefits greatly from the use of AD. Indeed, it gives us the possibility to modify the model of Eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered. This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
%
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data. One may notice how the system response follows the oscillations of the input signal.}
\end{figure}

>>>>

The methods we present are generic and applicable to a wide range of SDE models.
However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
The model equation is thus given by the SDE
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.

@@

The algorithm described above was implemented using the free and open-source programming language \emph{Julia} (version 0.3.9)~\cite{julia_web}. One of the attractive features of the language is its (fast growing) library of over 600 packages for numerical and scientific applications.
In our case, we made extensive use of the \emph{ReverseDiffSource} package, which provides a powerful tool for reverse-mode automated differentiation (AD) of arbitrarily complex user-defined expressions. Our algorithm benefits greatly from the use of AD. Indeed, it gives us the possibility to modify the model of Eq.~(\ref{sde}) and therefore the action~(\ref{action}) leaving the implementation of the algorithm unaltered. This clearly gives our program significant flexibility making it suitable for a much broader range of applications than the simple exemplary SDE model described here.
%
\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figs/FigRainData.png}
    \caption{\label{fig:rain_data_S}Sinusoidal input $r(t)$ (dashed line), "true" system realization with parameters $K_{\mbox{\small true}} = 50$ (in some arbitrary units of time) and $\gamma_{\mbox{\small true}} = 0.2$ (solid line) and synthetic observed data. One may notice how the system response follows the oscillations of the input signal.}
\end{figure}
